---
title: "DESeqAnalysis"
output: html_document
---

```{r setup, warning=FALSE, message=FALSE}
library(DESeq2)
library(magrittr)
library(EnhancedVolcano)
library(ashr)
library(fgsea)
library(plyr)
library(dplyr)
library(msigdbr)

dds4v5 = readRDS("outputs/dds4v5.rds")
dds4vM = readRDS("outputs/dds4vM.rds")
dds5vM = readRDS("outputs/dds5vM.rds")
```

# Analyzing the DESeq output

Here we run DESeq2 to get our DEGs for our contrasts of choice:
We will compare the three different timepoints against each other, generating a total of 3 contrasts.
1. protruding mouth vs day 4
2. day 4 vs day 5
3. protruding mouth vs day 5

```{r Generating DEGs for groups}
dds4v5$stageName <- droplevels(dds4v5$stageName)
dds4vM$stageName <- droplevels(dds4vM$stageName)
dds5vM$stageName <- droplevels(dds5vM$stageName)
dds4v5 = DESeq(dds4v5)
dds4vM = DESeq(dds4vM)
dds5vM = DESeq(dds5vM)
res4v5 <- results(dds4v5)
res4vM <- results(dds4vM)
res5vM <- results(dds5vM)
```

```{r outputs, warning=FALSE, message=FALSE}
res5vM = lfcShrink(dds=dds5vM, coef = "stageName_Day.5_vs_Protruding.mouth",res = res5vM, type = "ashr")
res4vM = lfcShrink(dds=dds4vM, coef = "stageName_Day.4_vs_Protruding.mouth",res = res4vM, type = "ashr")
res4v5 = lfcShrink(dds4v5, contrast = c("stageName", "Day.4", "Day.5"),res = res4v5, type="ashr")
```
Here we us LFCshrink via ashr to shrink our logfold2 changes from the DESeq outputs with respect to our contrasts. We do not use normal type as ashr and apeglm both outperform normal here.


## Dispersion Estimates

Here we glance into our data to see if the model is well fit to the data:
```{r dispestimates, fig.show='hold'}
plotDispEsts(dds4v5)
plotDispEsts(dds4vM)
plotDispEsts(dds5vM)
```

Everything looks mostly okay here; it mostly fits the "normal look" here.


## Top Genes expressed in Day 5 vs Mouth Protruding
```{r top counts 5vM, fig.show='hold', out.width="50%"}
(for (i in order(res5vM$padj, decreasing = F)[1:5]){
  plotCounts(dds5vM, gene=i, intgroup="stageName", main = "") 
  title(rowData(dds5vM)[rownames(res5vM[i,]),]$SYMBOL)
})
```

### Volcano Plot

```{r volcanofivevM, fig.height=16, fig.width=8}
fiveMplot = EnhancedVolcano(res5vM,
  lab = rowData(dds5vM)$SYMBOL,
  x = 'log2FoldChange',
  y = 'pvalue')

fiveMplot
ggsave("outputs/fivevsMvolcano.pdf", height = 16, width = 8)

```

## Top Genes for expressed Day4 vs mouth Protruding

```{r day4vM,fig.show='hold', out.width="50%"}
(for (i in order(res4vM$padj, decreasing = F)[1:5]){
  plotCounts(dds4vM, gene=i, intgroup="stageName", main = "") 
  title(rowData(dds4vM)[rownames(res4vM[i,]),]$SYMBOL)
})
```

### Volcano Plot

```{r volcanofourvM, fig.height=16, fig.width=8}
fourMplot = EnhancedVolcano(res4vM,
  lab = rowData(dds4vM)$SYMBOL,
  x = 'log2FoldChange',
  y = 'pvalue')

fourMplot
ggsave("outputs/fourvsMvolcano.pdf", height = 16, width = 8)
```

## significant genes 4v5

```{r day4v5,fig.show='hold', out.width="50%"}
(for (i in order(res4v5$padj, decreasing = F)[1:5]){
  plotCounts(dds4v5, gene=i, intgroup="stageName", main = "") 
  title(rowData(dds4v5)[rownames(res4v5[i,]),]$SYMBOL)
})
```

### Volcano plot 4v5

```{r volcanofourvfive, fig.height=16, fig.width=8}
fourfiveplot = EnhancedVolcano(res4v5,
  lab = rowData(dds4v5)$SYMBOL,
  x = 'log2FoldChange',
  y = 'pvalue')

fourfiveplot
ggsave("outputs/fourvsfivevolcano.pdf", height = 16, width = 8)
```
from all of this we do see that Hox genes are not the most DEgenes by counts. Instead we see the two DE genes that seem to be related to the trypsin? ctrl and prss1. 

Many of the other top genes are interspersed through many other gene ontologies, including fatty acid metaboloism/transport, and location specific markers i.e. (cyp3a65 is expressed in the membrane of these structures: digestive system, eye, gill, heart and ovary.)

## HOX genes only

Since this was my original question, let's filter and generate a heatmap of just the HOX genes, and let's see if we get some variability here.

```{r heatmapHOX, fig.height=12, fig.width=8}
library(pheatmap)
dds = readRDS("outputs/dds.rds")
ntd<-normTransform(dds)
select <- grep(rowData(ntd)$SYMBOL,pattern="hox")
df <- as.data.frame(colData(dds)[,"stageName"])
colnames(df)<-"stageName"
rownames(df)<-colnames(ntd)
pheatmap(assay(ntd)[select,], cluster_rows=TRUE, labels_row=grep(rowData(ntd)$SYMBOL,pattern="hox", value=TRUE),
         cluster_cols=TRUE, annotation_col=df)

```

From this we do see that unfortunately... all of the samples seem to just be relatively prolific in HOX expression across the three timepoints chosen. This may be due to the fact that these are later timepoints and are bulk samples. We capture the entire signal and as such we do not have the distinction of where these Hox genes are being expressed across the embryo, and we only know that the genes are being highly expressed. Still we do see some differences in the hox expression. Particularly, there does seem to be some differences in expression between the Protruding mouth when compared to day 4 and 5. 


### HOX gene exp plots
```{r HOXgene expression plots, fig.show='hold', out.width="30%", fig.width=20}
select <- grep(rowData(ntd)$SYMBOL,pattern="hox")
# get pvals for comparisons across each other
sig_4v5= (res4v5[select,]$padj < 0.05)
sig_4vM= (res4vM[select,]$padj < 0.05)
sig_5vM= (res5vM[select,]$padj < 0.05)
df_padj_vals = do.call(cbind, list(res4v5[select,]$padj, 
                     res4vM[select,]$padj, 
                     res5vM[select,]$padj))
# remove
remove_vals = -which(is.na(rowSums(df_padj_vals)))
select=select[remove_vals]
df_padj_vals=df_padj_vals[remove_vals,]
sig_4v5= sig_4v5[remove_vals]
sig_4vM= sig_4vM[remove_vals]
sig_5vM= sig_5vM[remove_vals]

(for (i in 1:length(select)){
  if (sig_4v5[i]==T | sig_4vM[i]==T | sig_5vM[i]==T){
    pvals=paste("pvals: 4v5:", round(df_padj_vals[i,1],3), 
                "4vM:", round(df_padj_vals[i,2],3),
                "5vM:", round(df_padj_vals[i,3],3))
    plotCounts(dds, gene=select[i], intgroup="stageName", main = "", sub=pvals) 
    title(rowData(dds)[rownames(dds[select[i],]),]$SYMBOL)
  }
})
```

Plotting out the HOX genes individually with their padj values we do see that we do get a better view that although we do not have significant pvalues for most hox genes when comparing the last two time points, HOX genes seem to trend downwards with respect to time for the three time points available.

```{r aspdf}
pdf("outputs/hox_gene_plots.pdf")
par(mfrow = c(2,1))
for (i in 1:length(select)){
  if (sig_4v5[i]==T | sig_4vM[i]==T | sig_5vM[i]==T){
    pvals=paste("pvals: 4v5:", round(df_padj_vals[i,1],3), 
                "4vM:", round(df_padj_vals[i,2],3),
                "5vM:", round(df_padj_vals[i,3],3))
    plotCounts(dds, gene=select[i], intgroup="stageName", main = "", sub=pvals) 
    title(rowData(dds)[rownames(dds[select[i],]),]$SYMBOL)
  }
}
dev.off()
```


## Pathway Analysis
```{r,echo=FALSE}
# for my own sanity
res4v5 <- results(dds4v5)
res4vM <- results(dds4vM)
res5vM <- results(dds5vM)
```

```{r fgsea, warning=FALSE, message=FALSE}
c2.cp7.gmt=msigdbr(species = "Danio rerio", category = "C2", subcategory = "CP:REACTOME")
msigdbr_list = split(x = c2.cp7.gmt$gene_symbol, f = c2.cp7.gmt$gs_name)

map= cbind(c2.cp7.gmt$ensembl_gene, c2.cp7.gmt$gene_symbol) %>% as.data.frame()
colnames(map) = c("ensembl_gene", "gene_symbol")
# here we clean up the names more so we can run fgsea
genes4v5 <- as.data.frame(res4v5)%>%
  mutate(ensembl_gene = rownames(.)) %>%
  join(., map, by = c("ensembl_gene"),
                 type="inner") %>%
  dplyr::select(gene_symbol, stat) %>% 
  stats::na.omit() %>% 
  dplyr::distinct() %>% 
  dplyr::group_by(gene_symbol) %>% 
  dplyr::summarize(stat=mean(stat)) %>% 
  with(., setNames(stat, gene_symbol))

genes4vM <- as.data.frame(res4vM)%>%
  mutate(ensembl_gene = rownames(.)) %>%
  join(., map, by = c("ensembl_gene"),
                 type="inner") %>%
  dplyr::select(gene_symbol, stat) %>% 
  stats::na.omit() %>% 
  dplyr::distinct() %>% 
  dplyr::group_by(gene_symbol) %>% 
  dplyr::summarize(stat=mean(stat)) %>% 
  with(., setNames(stat, gene_symbol))

genes5vM <- as.data.frame(res5vM)%>%
  mutate(ensembl_gene = rownames(.)) %>%
  join(., map, by = c("ensembl_gene"),
                 type="inner") %>%
  dplyr::select(gene_symbol, stat) %>% 
  stats::na.omit() %>% 
  dplyr::distinct() %>% 
  dplyr::group_by(gene_symbol) %>% 
  dplyr::summarize(stat=mean(stat)) %>% 
  with(., setNames(stat, gene_symbol))


out4vM = fgsea(pathways=msigdbr_list, stats = genes4vM, nperm=1000)
out5vM = fgsea(pathways=msigdbr_list, stats = genes5vM, nperm=1000)
out4v5 = fgsea(pathways=msigdbr_list, stats = genes4v5, nperm=1000)
```

```{r table 4vM}

out4vM %>% arrange(., desc(NES)) %>% .[,-c("leadingEdge", "nMoreExtreme")]
```

```{r table 5vM}

out5vM %>% arrange(.,desc(NES)) %>% .[,-c("leadingEdge", "nMoreExtreme")]
```


```{r table 4v5}

out4v5 %>% arrange(.,desc(NES)) %>% .[,-c("leadingEdge", "nMoreExtreme")]
```


## plots for GSEA

```{r gsea4vM, fig.show='hold', fig.height=12, fig.width=18}
# select the top 20 on both sides
out4vM_f <- slice_max(out4vM, n=20, order_by = NES) %>% 
  rbind(., slice_min(out4vM, n = 20, order_by = NES))
ggplot(out4vM_f, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title="Top 20 C2 Reactome Pathways NES from GSEA: Day4vsMouth") + 
  theme_minimal()

ggsave("outputs/GSEA4vM.pdf", height = 16, width = 8)

```

```{r gsea5vM, fig.show='hold', fig.height=12, fig.width=18}
# select the top 20 on both sides
out5vM_f <- slice_max(out5vM, n=20, order_by = NES) %>% 
  rbind(., slice_min(out5vM, n = 20, order_by = NES))
ggplot(out5vM_f, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title="Top 20 C2 Reactome Pathways NES from GSEA: Day5vsMouth") + 
  theme_minimal()

ggsave("outputs/GSEA5vM.pdf", height = 16, width = 8)
```


```{r gsea4v5, fig.show='hold', fig.height=12, fig.width=18}
# select the top 20 on both sides
out4v5_f <- slice_max(out4v5, n=20, order_by = NES) %>% 
  rbind(., slice_min(out4v5, n = 20, order_by = NES))
ggplot(out4v5_f, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title="Top 20 C2 Reactome Pathways NES from GSEA: Day4vsDay5") + 
  theme_minimal()

ggsave("outputs/GSEA4v5.pdf", height = 16, width = 8)
```

Some variability in the pathways that were enriched, but we do have decent number of similarities for the 
4vM and 5vM contrasts.


```{r exploring gene overlaps across Gene Enrichments}
out4vM_f$leadingEdge %>% unlist() %>% table() %>% sort()
out5vM_f$leadingEdge %>% unlist() %>% table() %>% sort()
out4v5_f$leadingEdge %>% unlist() %>% table() %>% sort()
```

```{r heatmapribo, fig.height=16, fig.width=8}
selectr <- grep(rowData(ntd)$SYMBOL,pattern="^rps[0-9]|^rpl[0-9]")
pheatmap(assay(ntd)[selectr,], cluster_rows=TRUE, 
         labels_row=grep(rowData(ntd)$SYMBOL,
                         pattern="^rps[0-9]|^rpl[0-9]", value=TRUE),
         cluster_cols=TRUE, annotation_col=df)
```